---
title: "EDS-Cancer-EQI"
author: "Uma Bhat"
date: "2024-03-25"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidycensus)
```

Cancer instance data was obtained from NCDHHS in PDF form and was converted to .csv format using Tabula.

```{r data}
eqi_2018 <- read_csv("/Users/umabhat/Documents/UNC 2023-2024/STOR 320/Exploratory Data Analysis/Eqidata_all_domains_non-transformed_2013JULY18 (1).csv", show_col_types = FALSE)
cancer_2018 <- read_csv("/Users/umabhat/Documents/UNC 2023-2024/STOR 320/EDA/instance-data.csv", show_col_types = FALSE)
```

```{r nc-specific}
eqi_2018 <- eqi_2018 %>% 
  filter(STATE == "NC")
```

For purposes of this study, we will limit the variables we address to:

`pct_unemp` - percent unemployed

`med_hh_inc` - median household income 

`A_PCBs` - Polychlorinated biphenyls in air (carcinogenic)

These variables are specific to just a few key social determinants of health and known carcinogens. 

```{r cleaning}
eqi_2018 %>%
  select(
    `pct_unemp`,
    `med_hh_inc`,
    `A_PCBs`,
  )
```

Which counties saw the highest cancer instance rates? 

Which counties had the poorest EQI levels for our selected variables?

*To expand this study, we'd try to find the association between EQI levels over time with cancer instance rates over time.*

Has the number of instances of cancer increased in certain counties' populations over time? (Plot)

Have certain kinds of cancers been more prevalent in the past five years? 

How closely linked are EQI levels and cancer instance rates? Based on EQI levels, could you predict what the cancer instance rate will be in a county?




